Semi-supervised low-rank logistic regression for high-dimensional neuroimaging data

### 2015-05-18 paper_nips/paper.tex
add missing files

@@

\author{Danilo Bzdok, Michael Eickenberg, Olivier, Grisel,
  Bertrand Thirion,
  Ga\"el Varoquaux \\\\\textbf{\textit{email:} }firstname.lastname@inria.fr}

@@

\linebreak

@@

Unsupervised and supervised learning were combined into a low-rank logistic
regression problem. The 13,657 z values from each activity map were
subject to a first linear
projection into Z latent components (i.e., 1, 5, 10, .., 100).
These hidden brain networks
loadings were subsequently projected into the 18 class space for multinomial
logistic regression. The goal was to find the two weight matrices
(input-to-hidden: 13,657 x Z and hidden-to-output: Z x 18) and their
corresponding bias vectors. Non-linearities were not applied on the
transformation results. Weights and biases were initialized by Gaussian
random noise. Gradient descent updated these matrices and vectors in each
iteration (100 samples per batch, 250 epochs). Using the chain rule, the
partial
derivates for the update were computed for the transformation into the
latent space and the subsequent transformation into the class space. We choose
the RMSprop algorithm \cite{dauphin15} with an initial learning rate of
0.01. Early stopping ensured that the best weight matrices/vectors were
retained in each epoch cycle. This was evaluated by the prediction
accuracy on a validation set (10\% of the training data) at each iteration.
\linebreak

@@

FORMULA
X e R^d real-valued input
'reconstruction error' criterion
encoder: input -> hidden reprsentation
Its parameter set is θ = {W, b}, where W is a d′ × d weight matrix and b is an offset vector of dimensionality d′.
decoder: gθ′ (y) = s(W′y + b′), (3)
with appropriately sized parameters θ′ = {W′,b′}.
where b0 and b1 are biase vectors
average reconstruction error
pen. avoids learning trivial soutoin - identitiy

coompute the error derivates of the weights
-> gradually modified according to possible compression algorithms

affine autoencoder with squared error loss
tied weights beteen W0 and W1, i.e., W1 = W0^T
X hat is a dterministic function of X
It can thus be said that training an autoencoder to minimize reconstruction error amounts to maximizing a lower bound on the mutual information between input X and learnt repre- sentation Y .

rest data is already compressed to a sparse PCA -> remote form of 
pretraining of the first layer
bottleneck -> unter-complete representing, i.e. lossy compression

@@

HINTS:
In fact, the constraint by a rest-data autoencoder qualifies as a
\textit{hint} rather than regularization \cite{abu1994hints}.
Its purpose is not to prevent overfitting but to introduce
prior knowledge on known properties of the unknown target function $f$.
Rather than relying only on input-output pairs in the learning process,
we thus narrow our hypothesis set to the biologically most plausible solutions.
That is, we reduce the search space in a way that
is compatible with the expected representation of BOLD activity signals.

@@

{arg\,min}_\theta \hspace{0.2cm} {\mathcal{L}}(\theta, \lambda) = \lambda\frac{1}{N_{X_{task}}} \sum_{i=0}^{N_{X_{task}}} \log(P(Y=y^{(i)}|x^{(i)}; W,b)) \hspace{0.2cm} + \hspace{0.2cm} (1-\lambda)\frac{1}{N_{X_{rest}}}\|X_{rest}-\hat X_{rest}(W,b)\|^2

mutiple output units

\linebreak

As solver, we chose \textit{rmsprop},
a mini-batch version of rprop \cite{dauphin2015rmsprop}.
This procedure dictates an adaptive learning rate
for each model parameter by
scaled gradients from a running average.
Gradient normalization by RMSprop
is known to effectively exploit curvature information.
We opted for a small batch size of $100$, given the high degree of
redundancy in $X_{rest}$ and $X_{task}$.
The matrix parameters were initalized by Gaussian random values multiplied
by a gain of $0.004$. The bias parameters were initalized to be zero.
The normalization factor is computed by

\begin{eqnarray*}
v^{(t+1)} = \rho v^{(t+1)} + (1 - \rho)(\nabla f)^2
\end{eqnarray*}

where $0 < \rho < 1$ constitutes the decay rate. $\rho$ was set to
$0.9$ to deemphasize the magnitude of the gradient.
The update step is
given by

\begin{eqnarray*}
\theta^{(t+1)} = \theta^{(t)} + \alpha \frac{\nabla f(\theta^{(t)})}{\sqrt{v^{(t+1)} + \epsilon}}
\end{eqnarray*}

where $\alpha$ is the learning rate and $\epsilon$ a damping factor.
The hyper-parameter $\alpha$ was set to $0.00001$ by prior manuel
cross-validation and $\epsilon$ was set to $1^{-6}$.
%
Note that we have also experimented with other solvers
(stochastic gradient descent, adadelta, and adagrad) but found that
rmsprop converged faster and with higher generalization performance.

\linebreak

@@

theano/python
The analyses were performed in Python.
We used \textit{nilearn} to handle
the high-dimensional neuroimaging data and
\textit{theano} for automatic gradient
differentiation of symbolic computation graphs.
All Python scripts that generated the results are
accessible online for reproducibility and reuse
\url{http://github.com/banilo/nips2015}.

@@

statistical structure^

@@

>>>>

\author{Danilo Bzdok, Michael Eickenberg, Olivier Grisel,
  Bertrand Thirion,
  Ga\"el Varoquaux \\\\\textbf{\textit{email:} }firstname.lastname@inria.fr}

@@

@@

@@

Unsupervised and supervised learning were combined into a low-rank logistic
regression problem. The 13,657 z values from each activity map were
subject to a first linear
projection into Z latent components (i.e., 1, 5, 10, .., 100).
These hidden brain networks
loadings were subsequently projected into the 18 class space for multinomial
logistic regression. The goal was to find the two weight matrices
(input-to-hidden: 13,657 x Z and hidden-to-output: Z x 18) and their
corresponding bias vectors. Non-linearities were not applied on the
transformation results. Weights and biases were initialized by Gaussian
random noise. Gradient descent updated these matrices and vectors in each
iteration (100 samples per batch, 250 epochs). Using the chain rule, the
partial
derivates for the update were computed for the transformation into the
latent space and the subsequent transformation into the class space. We choose
the RMSprop algorithm \cite{rmsprop} with an initial learning rate of
0.01. Early stopping ensured that the best weight matrices/vectors were
retained in each epoch cycle. This was evaluated by the prediction
accuracy on a validation set (10\% of the training data) at each iteration.
\linebreak

@@

FORMULA
%X e R^d real-valued input
'reconstruction error' criterion
encoder: input -> hidden reprsentation
%Its parameter set is θ = {W, b}, where W is a d′ × d weight matrix and b is an offset vector of dimensionality d′.
%decoder: gθ′ (y) = s(W′y + b′), (3)
%with appropriately sized parameters θ′ = {W′,b′}.
where b0 and b1 are biase vectors
average reconstruction error
pen. avoids learning trivial soutoin - identitiy

coompute the error derivates of the weights
gradually modified according to possible compression algorithms

affine autoencoder with squared error loss
%tied weights beteen W0 and W1, i.e., W1 = W0^T
X hat is a dterministic function of X
It can thus be said that training an autoencoder to minimize reconstruction error amounts to maximizing a lower bound on the mutual information between input X and learnt repre- sentation Y .

rest data is already compressed to a sparse PCA remote form of 
pretraining of the first layer
bottleneck -> unter-complete representing, i.e. lossy compression

@@

\paragraph{Hints.}
In fact, the constraint by a rest-data autoencoder qualifies as a
\textit{hint} rather than regularization \cite{abu1994hints}.
Its purpose is not to prevent overfitting but to introduce
prior knowledge on known properties of the unknown target function $f$.
Rather than relying only on input-output pairs in the learning process,
we thus narrow our hypothesis set to the biologically most plausible solutions.
That is, we reduce the search space in a way that
is compatible with the expected representation of BOLD activity signals.

@@

\begin{eqnarray}
  \begin{split}
{arg\,min}_\theta \hspace{0.2cm} {\mathcal{L}}(\theta, \lambda) &= \lambda\frac{1}{N_{X_{task}}} \sum_{i=0}^{N_{X_{task}}} \log(P(Y=y^{(i)}|x^{(i)}; W,b)) \hspace{0.2cm} \\
&+ \hspace{0.2cm} (1-\lambda)\frac{1}{N_{X_{rest}}}\|X_{rest}-\hat X_{rest}(W,b)\|^2
  \end{split}
  \label{eq:loss_equ}
\end{eqnarray}

mutiple output units
aCcording to equation \eqref{eq:loss_equ}
As solver, we chose \textit{rmsprop} \cite{rmsprop},
a mini-batch version of rprop.
This procedure dictates an adaptive learning rate
for each model parameter by
scaled gradients from a running average.
Gradient normalization by RMSprop
is known to effectively exploit curvature information.
We opted for a small batch size of $100$, given the high degree of
redundancy in $X_{rest}$ and $X_{task}$.
The matrix parameters were initalized by Gaussian random values multiplied
by a gain of $0.004$. The bias parameters were initalized to be zero.

With a slight abuse of notation, let $\theta$ denote a component of $\theta$.
The normalization factor and the update rule for $\theta$
are given by

\begin{eqnarray}
  \begin{split}
v^{(t+1)} &= \rho v^{(t+1)} + (1 - \rho)\left(\frac{\partial f}{\partial \theta}\right)^2
%v^{(t+1)} = \rho v^{(t+1)} + (1 - \rho)˜|\nabla f
\\
\theta^{(t+1)} &= \theta^{(t)} + \alpha \frac{\nabla f(\theta^{(t)})}{\sqrt{v^{(t+1)} + \epsilon}},
  \end{split}
\end{eqnarray}

where $0 < \rho < 1$ constitutes the decay rate. $\rho$ was set to
$0.9$ to deemphasize the magnitude of the gradient.
Further $\alpha$ is the learning rate and $\epsilon$ a global damping factor.
The hyper-parameter $\alpha$ was set to $0.00001$ by prior manuel
cross-validation and $\epsilon$ was set to $1^{-6}$.
%
Note that we have also experimented with other solvers
(stochastic gradient descent, adadelta, and adagrad) but found that
rmsprop converged faster and with higher generalization performance.

@@

theanopython
The analyses were performed in Python.
We used \textit{nilearn} to handle
the high-dimensional neuroimaging data 
\cite{abrah14}
and
\textit{theano} for automatic
differentiation of symbolic computation graphs
\cite{bastien2012theano, bergstra2010theano}.
All Python scripts that generated the results are
accessible online for reproducibility and reuse
\url{http://github.com/banilo/nips2015}.

@@

statistical structure

@@

FUTURE
link the restings-tate component to the pattern sof cognitive proceses
reduce information von resting-state data in neurological and psychiatric
populations for discovery of neurobiological sub-groups as well as
prediction of disease trajectories and drug responses
